Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae

### 2020-11-09 introduction.tex

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network (Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}) within $5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN. \tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

@@

Using results from 2D CCSN numerical simulations as a background with respect to which to solve an eigenvalue problem~\citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). \tf{This last sentence might be improved.}
This reveals that it is posible to perform CCSN asteroseismology {under realistic conditions} and serves as a starting point to carry out inference of astrophysical parameters of PNSs. 
Further work was presented in {\citep{Torres:2019b} who found that it is possible to derive simple relations} between the instantaneous frequency of the g-mode and the mass and radius of the PNS {at each time of the numerical evolutions}. These relations are universal as they do not depend {on the equation of state (EOS) or on the mass of the progenitor {and they only depend weakly on} the numerical code used {(see discussion in Section~\ref{sec:simulations})}. {Similar universal relations have been discussed by \citep{Sotani:2020a,Sotani:2020b} who also found that they do not depend on the dimensionality (1D, 2D or 3D) of the numerical simulation used as a background.

>>>>

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}, within $\sim 5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN. \tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.}\mab{MAB:I'm in favor of deleting it as it is rather obvious we are considering neutrino driven CCSN} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

@@

Using results from 2D CCSN numerical simulations as a background with respect to which to solve an eigenvalue problem~\citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). \tf{This last sentence might be improved.} \mab{MAB: What about: Computing the mode spectrum of the region within the shock (including the PNS and the post-shock region) consists in solving an eigenvalue problem using background simulations. Using results from 2D CCSN numerical simulations \citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). }

This reveals that it is posible to perform CCSN asteroseismology {under realistic conditions} and serves as a starting point to carry out inference of astrophysical parameters of PNSs. 
Further work was presented in {\citep{Torres:2019b} who found that it is possible to derive simple relations} between the instantaneous frequency of the g-mode and the mass and radius of the PNS {at each time of the numerical evolutions}. These relations are universal as they do not depend {on the equation of state (EOS) or on the mass of the progenitor {and they only depend weakly on} the numerical code used {(see discussion in Section~\ref{sec:simulations})}. {Similar universal relations have been discussed by \citep{Sotani:2020a,Sotani:2020b} who also found that they do not depend on the dimensionality (1D, 2D or 3D) of the numerical simulation used as a background.

### 2020-11-09 results.tex

 \caption{Boxplots of the $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realizations are considered for each distance. \tf{We should explain a bit more the symbols and colours that appear in the plots.}}
  \label{fig:s20results}
\end{figure}

@@

For each of the noise realizations we reconstruct the ratio time series {$r_i$} of length $N$ starting from the left side of the spectrogram and constraining the beginning of the track to be smaller than \unit[200]{Hz}. \tf{Perhaps we should mention that this value is chosen using the information on the initial mode frequency from the simulations.}
The reconstructed ratio is then compared to the ``true'' ratio {$r_i^0$} derived from the PNS mass and radius computed from the {\texttt s20S} simulation. The top panel of Fig.~\ref{fig:s20results} shows the distribution of the fraction of the ratio values {$r_i^0$} that fall within the 95\% confidence interval of {$r_i$} as a function of the distance to the source. This quantity, hence, gives information about the {\it coverage} of the reconstructed ratio. The coverage takes maximum values when the source is located within a few kpc and then decreases with the distance.

To better quantify how well we reconstruct the ratio we also consider  the mean of the relative error of $r_i$ along 
the track of the spectrogram, $\Delta$,  
\begin{equation}
\Delta=\frac{1}{N}\sum_1^N\frac{|r_i-r_i^0|}{r_i^0}\,.
\end{equation}
The values of $\Delta$ for each of the 100 noise realizations are shown as a function of the distance
in the bottom panel of Fig.~\ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%. At closer distances $\Delta$ is smaller but it does not vanish. This reflects the approximate nature of the model used for $r$. It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision up to distances of $\sim$ 9 kpc (for this particular waveform) with a coverage value
larger than 80\%. We note that there are a few noise realizations for distances below 9 kpc for which
$\Delta$ takes large values, indicating that the method fails to accurately reconstruct the ratio in those case. \tf{Do we know why that is the case? What is intrinsically different from one noise realisation to another that makes the method more prone to failure? I think we should provide some explanation here, if we have it.}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_coverage_allwvfs}
 \caption{Median of the coverage for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the  distance to the source. The ``no signal'' line and band show the median and first and third quartile of coverage in absence of any signal. The line of the latter is zero and overlaps with the horizontal axis (see main text for details). \tf{Perhaps is more consistent to use in the legend the same model names as in Table 1. Some of the colours are really hard to see, particularly for s20S. That could perhaps be improved if we use thicker lines. I'm confused about the quartiles shown.}} \label{fig:aLIGO_cov_allwvf}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_delta_allwvfs}
 \caption{Same as Fig.~\ref{fig:aLIGO_cov_allwvf} but for the relative error $\Delta$. \tf{Same comments as for Fig.~\ref{fig:aLIGO_cov_allwvf}.}}
% Median of $\Delta$ for 8 CCSN waveforms embedded in aLIGO noise and located at different distance from the Earth. The ``no signal'' line and band show the median and first and third quartile of $\Delta$ in absence of any signal.} 
\label{fig:aLIGO_prec_allwvf}
\end{figure}

We have tested that the method does not depend on the specific features of the waveform of model {\texttt s20S} by repeating the procedure for the remaining seven waveforms of the {\it test set} described in Section \ref{sec:simulations} covering
a large range of progenitor masses. Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from model {\texttt s11} and to a lesser extent model {\texttt s20S}, the ratio is well reconstructed for all waveforms up to a distance of $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio we have run 100 simulations without injecting a signal and have measured the corresponding coverage for the reconstructed ratios.
The median of the coverage as well as the 95 quartile are shown in Figure \ref{fig:aLIGO_cov_allwvf}. \tf{This seems inconsistent with what is written in the caption, which indicates the third quartile.}
The noise only median value is identically zero in this case. However, note that it could be different from zero because
the g-mode reconstruction algorithm is looking for a continuously  increasing frequency track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effect explains why outliers can reach
values as high as 80\%. \tf{I don't understand this last comment.}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for all GW detectors analyzed in this study. The values are shown for all eight CCSN waveforms of our sample assuming that the source is optimally oriented with respect to the detector. \tf{I guess that for CE2 s15G is overlapping with s25. Should we mention this in the caption? Use AdV instead of ADV?}} 
\label{fig:distances}
\end{figure}

>>>>

 \caption{Boxplots of the $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realizations are considered for each distance. The orange lines indicate the median values while the empty rectangles indicate the first and third quartiles. The ``+'' markers are outliers outside the first and third quartiles.}
  \label{fig:s20results}
\end{figure}

@@

For each of the noise realizations we reconstruct the ratio time series {$r_i$} of length $N$ starting from the left side of the spectrogram and constraining the beginning of the track to be smaller than \unit[200]{Hz}. \mab{This value is chosen using the information on the initial mode frequency from the simulations.}
The reconstructed ratio is then compared to the ``true'' ratio {$r_i^0$} derived from the PNS mass and radius computed from the {\texttt s20S} simulation. The top panel of Fig.~\ref{fig:s20results} shows the distribution of the fraction of the ratio values {$r_i^0$} that fall within the 95\% confidence interval of {$r_i$} as a function of the distance to the source. This quantity, hence, gives information about the {\it coverage} of the reconstructed ratio. The coverage takes maximum values when the source is located within a few kpc and then decreases with the distance.

To better quantify how well we reconstruct the ratio we also consider  the mean of the relative error of $r_i$ along 
the track of the spectrogram, $\Delta$,  
\begin{equation}
\Delta=\frac{1}{N}\sum_1^N\frac{|r_i-r_i^0|}{r_i^0}\,.
\end{equation}
The values of $\Delta$ for each of the 100 noise realizations are shown as a function of the distance
in the bottom panel of Fig.~\ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%. At closer distances $\Delta$ is smaller but it does not vanish. This reflects the approximate nature of the model used for $r$. It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision up to distances of $\sim$ 9 kpc (for this particular waveform) with a coverage value
larger than 80\%. We note that there are a few noise realizations for distances below 9 kpc for which
$\Delta$ takes large values, indicating that the method fails to accurately reconstruct the ratio in those cases. \mab{The main cause of failure is the split of the ridge in few blobs when the signal becomes weaker and weaker.}
%\tf{Do we know why that is the case? What is intrinsically different from one noise realisation to another that makes the method more prone to failure? I think we should provide some explanation here, if we have it.}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_coverage_allwvfs}
 \caption{Median of the coverage for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the  distance to the source. \mab{The ``no signal'' line shows the median of coverage in absence of any signal. In this example, the median is null and overlaps with the horizontal axis. The blue band boundaries are given by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles of coverage in absence of any signal.} \tf{Perhaps is more consistent to use in the legend the same model names as in Table 1. Some of the colours are really hard to see, particularly for s20S. That could perhaps be improved if we use thicker lines. I'm confused about the quartiles shown.}} \label{fig:aLIGO_cov_allwvf}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_delta_allwvfs}
 \caption{Same as Fig.~\ref{fig:aLIGO_cov_allwvf} but for the relative error $\Delta$.}
% Median of $\Delta$ for 8 CCSN waveforms embedded in aLIGO noise and located at different distance from the Earth. The ``no signal'' line and band show the median and first and third quartile of $\Delta$ in absence of any signal.} 
\label{fig:aLIGO_prec_allwvf}
\end{figure}

We have tested that the method does not depend on the specific features of the waveform of model {\texttt s20S} by repeating the procedure for the remaining seven waveforms of the {\it test set} described in Section \ref{sec:simulations} covering
a large range of progenitor masses. Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from model {\texttt s11} and to a lesser extent model {\texttt s20S}, the ratio is well reconstructed for all waveforms up to a distance of $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio we have run 100 simulations without injecting a signal and have measured the corresponding coverage for the reconstructed ratios.
\mab{The median of the coverage as well as the band defined by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles are shown in Figure \ref{fig:aLIGO_cov_allwvf}.}
The noise only median value is identically zero in this case. However, note that it could be different
from zero because
the g-mode reconstruction algorithm is looking for a continuously  increasing frequency track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effect explains why certain values of overlap can reach
values as high as 80\% even when no signal is added to the noise. \tf{I don't understand this last comment.} \mab{Toni: is that clearer?}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for all GW detectors analyzed in this study. The values are shown for all eight CCSN waveforms of our sample assuming that the source is optimally oriented with respect to the detector. Some waveform maximal distance markers are not visible because the do overlap (for instance \texttt{s15G} is overlapping with \texttt{s25} for CE2). \tf{Use AdV instead of ADV?}} 
\label{fig:distances}
\end{figure}
